---
output: html_document
---
### Miles per Gallon (MPG) on Manual vs Automatic Transmission Vehicles
#### Frank D. Evans - Johns Hopkins Data Science, Regression Modeling



### Executive Summary




### Exploratory Analysis

Load the data set, and required libraries.
```{r}
library(datasets)
library(MASS)
data(mtcars)
```

The key questions of interest concern the car mileage (mpg) and type of transmission 
(0:automatic, 1:manual).  Based on data documentation, it can safely be assumed that 
transmission type is a discrete factor dimension, while mpg is measured on a bounded,
yet continuous scale that will consist exclusively of positive values.  To verify
assumptions and get a starting point of scale of the data, we will call a summary of
the mpg attribute.
```{r}
summary(mtcars$mpg)
```

To get a sense of scale, we'll examine the number of records by factor for the transmission
variable.
```{r}
table(mtcars$am)
```

To get a prelimiary sense of data difference, we can run a summary separated by the 
transmission factor.
```{r}
by(mtcars$mpg, mtcars$am, summary)
```

In addition to quartiles, it is useful to get a feel for the distribution of the mpg
data with standard deviations at a 95% confidence interval.
```{r}
c(avg = mean(mtcars$mpg), 
  sigma = sd(mtcars$mpg), 
  sig_hi = mean(mtcars$mpg) + 2 * (sd(mtcars$mpg)),
  sig_lo = mean(mtcars$mpg) - 2 * (sd(mtcars$mpg)))
```

Next we can examine the straight correlation of the two factors of interest, which will
indeed verify that there does exist a relationship between the variables of interest 
independent of corroborating factors.
```{r}
cor(mtcars$mpg, mtcars$am)
```

### Regression Analysis
#### Single Variable
We first want to test a relationship directly between the variables of interest.  A
single variable regression seems to imply a considerable relationship.
```{r}
fit_1var <- lm(mtcars$mpg ~ mtcars$am)
summary(fit_1var)
```
The coefficient of 7.2 would imply that in the abscence of other factors, it should 
be expected that a manual tranmission car (encoded in the am variable as 1) attributes 
to an increase of 7.2 mpg.  However, with a t value of 4.1 and a considerable band of
variation around the residuals--this model appears to be of poor fit.  Further investigation
of the plot of the residuals (Appendix: Figure 1) confirms the probable poor fit of the
single variable model.

#### Multi-Variable Analysis
Next, we consider all of the factors from the dataset.  A pairs plot of the non-derivative 
factors in the data show there to be likely considerable statistical relationships among 
several key factors throughout the data (Appendix: Figure 3).
```{r}
fit_multi <- lm(mpg ~ . , data = mtcars)
summary(fit_multi)
```
The estimate coefficient of transmission type against mpg falls considerably to 2.5 when all
factors are considered.  However, the very high p-value of 0.23 indicates that this has a 
relatively high probability of not being significant.  

To find a better fitting model that best captures the variability given the variables given,
a generalized linear model is fit against the principle variables. (Note: factor dimensional
variables are fed into the model with a factor object conversion to automatically generate
the necessary dummy variables.)  Then, we leverage a stepwise algorithm to consider alternate
multi-variate regression models for better candidates.
```{r results='hide'}
step_fit <- glm(mpg ~ as.factor(am) + as.factor(cyl) + as.factor(gear) + disp + hp + drat + wt, data = mtcars)
model_fit <- stepAIC(step_fit, direction = 'both')
```
```{r}
model_fit$anova
```
The results of an ANOVA test from the stepwise fit algorithm suggest a final model based on
the transmission type compunded with the number of cylinders in the car's engine (cyl), the
horsepower of the car (hp) and the weight of the car (wt).  From an intuition perspective, this
makes probable sense.

A final model is built from these factors.
```{r}
final_model <- glm(mpg ~ as.factor(am) + as.factor(cyl) + hp + wt, data = mtcars)
summary(final_model)
```
The results of the sugested model suggest a much lower relationship, with a manual transmission
attributable to a 1.8 mile per gallon lift when controlled for key corroborating factors.
However, the variability of this measure is still present as the high p-value of 0.21 for the
transmission factor means that there is a relatively low probability that the relationship as
observed in this data is significant.

More significant factors are found among the engine size, the horsepower of the engine, and the
weight of the car.  The weight of the car being the most sginificant factor statistically, and
interpreted as a probable 2.5 mile per gallon decrease with every 1,000 pounds of additional
weight in the car itself.  Behind this factor, is a likely significant factor that the presence
of a 6 cylendar engine as compared to a 4 cylandar baseline attributes a reduction of mpg by
3.0 miles per gallon.

A major corroborating factor surrounding statistical signifigance in this case is the high
ratio of the total number of available variables (11) to the number fo observations (32).
The resultant model is likely to improve considerably with the addition of more data.  However,
an examination of the model fit and residual fit (Appendix: Figure 4) show that given the data
and variables for analysis, incorporating the additional variable results in a significantly
better explantory and predictive model.

## Appendix

### Figure 1: Single Variable Regression Plot and Residuals
```{r fig.height=5, fig.width=8}
par(mfrow = c(1,2))
plot(mtcars$am, mtcars$mpg, main = '1 Variable Regression: MPG vs Transmission')
abline(fit_1var)
plot(mtcars$am, resid(fit_1var), main = 'Residuals: 1 Variable Exploratory')
abline(h = 0)
```

### Figure 2: Single Variable Regression - Full Model Fit Plots
```{r fig.height=8, fig.width=8}
layout(matrix(c(1,2,3,4),2,2))
plot(fit_1var, main = 'Single Variable: mpg vs transmission')
```

### Figure 3: Pairs Plot for MTCars Attributes
```{r fig.height=5, fig.width=5}
test_vars <- c('mpg','cyl','disp','hp','wt','gear','carb')
pairs(x = mtcars[,test_vars], 
      panel = panel.smooth,
      main = 'Pairs Plot with Smoothing for MTCars Attributes')
```

### Figure 4: Final Model Regression - Full Model Fit Plots
```{r fig.height=8, fig.width=8}
layout(matrix(c(1,2,3,4),2,2))
plot(final_model, main = 'Final Model: mpg vs transmission and other factors')
```
