---
output: html_document
---
#### Miles per Gallon (MPG) on Manual vs Automatic Transmission Vehicles
*Frank D. Evans - Johns Hopkins Data Science, Regression Modeling*  

**Executive Summary**  
The key questions of interest concern the car mileage (mpg) and type of transmission 
(0:automatic, 1:manual).  Considering the relationship only between those factors
shows a strong relationship--though one with a low probability of being statistically
signicant given the size of the data.  When the other key attributes of the car are 
held constant, the relationship drops strongly to the transmission type--and other 
factors about the car show a high probability of being more strongly related to the
mileage.

**Exploratory Analysis**  
Load the data set, and required libraries.
```{r}
library(datasets); library(MASS); data(mtcars)
```
Based on data documentation, it can safely be assumed that transmission type is a discrete 
factor dimension, while mpg is measured on a bounded, yet continuous scale that will consist 
exclusively of positive values.  To verify assumptions and get a starting point of scale of 
the data, we will call a summary of the mpg attribute separated by the transmission factor.
```{r}
by(mtcars$mpg, mtcars$am, summary)
```
To get a sense of scale, we'll examine the number of records by factor for the transmission
variable.
```{r}
c(sum(mtcars$am == 0), sum(mtcars$am == 1))
```
#### Regression Analysis
**Single Variable**  
We first want to test a relationship directly between the variables of interest.  A
single variable regression seems to imply a considerable relationship.
```{r}
fit_1var <- lm(mtcars$mpg ~ mtcars$am)
summary(fit_1var)$coefficients
```
The coefficient of 7.2 would imply that a manual transmission should expect an mpg increase of
7.2 miles per gallon.  The t value and residual variation implies the model is a poor fit,
confirmed by the plor of the residuals (Appendix: Figure 1).

**Multi-Variable Analysis**
To find a better fitting model a generalized linear model is fit against the principle variables. 
Dimensional variables are fed into the model as a factor to generate dummy variables, and we 
use a stepwise method to consider alternate multi-variate regression models for better candidates.
```{r results='hide'}
step_fit <- glm(mpg ~ as.factor(am) + as.factor(cyl) + as.factor(gear) + disp + hp + drat + wt, data = mtcars)
model_fit <- stepAIC(step_fit, direction = 'both'); model_fit$anova
```
The results of an ANOVA test from the stepwise fit algorithm suggest a final model based on
the transmission type compunded with the number of cylinders in the car's engine (cyl), the
horsepower of the car (hp) and the weight of the car (wt).  From an intuition perspective, this
makes probable sense.
```{r}
final_model <- glm(mpg ~ as.factor(am) + as.factor(cyl) + hp + wt, data = mtcars)
summary(final_model)$coefficients; summary(final_model$residuals)
```
The results of the sugested model suggest a much lower relationship, with a manual transmission
attributable to a 1.8 mile per gallon lift when controlled for key corroborating factors.
However, the high p-value of 0.21 for the transmission factor means that there is a relatively 
low probability that the relationship as observed in this data is significant.

More significant factors are found among the engine size, horsepower, and the weight of the car.
Behind this factor, is a likely significant factor that the presence of a 6 cylindar engine as 
compared to a 4 cylindar baseline attributes a reduction of mpg by 3.0 miles per gallon.

A major corroborating factor surrounding statistical signifigance in this case is the high
ratio of the total number of available variables (11) to the number of observations (32).
Examination of the model fit and residual fit (Appendix: Figure 4) show that given the data
and variables for analysis, incorporating the additional variable results in a significantly
better explantory and predictive model.

#### Appendix

**Figure 1: Single Variable Regression Plot and Residuals**  
```{r fig.height=4, fig.width=8}
par(mfrow = c(1,2))
plot(mtcars$am, mtcars$mpg, main = '1 Variable Regression: MPG vs Trans')
abline(fit_1var)
plot(mtcars$am, resid(fit_1var), main = 'Residuals: 1 Variable Exploratory')
abline(h = 0)
```

**Figure 2: Single Variable Regression - Full Model Fit Plots**  
```{r fig.height=5, fig.width=8}
layout(matrix(c(1,2,3,4),2,2))
plot(fit_1var)
```

**Figure 3: Pairs Plot for MTCars Attributes**  
```{r fig.height=4, fig.width=4}
test_vars <- c('mpg','cyl','disp','hp','wt','gear','carb')
pairs(x = mtcars[,test_vars], 
      panel = panel.smooth,
      main = 'MTCars Attributes')
```

**Figure 4: Final Model Regression - Full Model Fit Plots**  
```{r fig.height=4, fig.width=4}
layout(matrix(c(1,2,3,4),2,2))
plot(final_model)
```
