Practical Machine Learning - Prediction Assignment Writeup
==========================================================

For this assignment I analyzed the provided data to determine what activity an individual perform.
To do this I made use of caret and randomForest, this allowed me to generate correct answers for
each of the 20 test data cases provided in this assignment.  I made use of a seed value for 
consistent results.


```{r}
library(Hmisc)
library(caret)
library(randomForest)
library(foreach)
library(doParallel)
set.seed(2048)
options(warn=-1)
```

First, I loaded the data both from the provided training and test data provided by COURSERA.
Some values contained a "#DIV/0!" that I replaced with an NA value.

```{r}
training_data <- read.csv("pml-training.csv", na.strings=c("#DIV/0!") )
evaluation_data <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!") )
```

I also casted all columns 8 to the end to be numeric.

```{r}
for(i in c(8:ncol(training_data)-1)) {training_data[,i] = as.numeric(as.character(training_data[,i]))}

for(i in c(8:ncol(evaluation_data)-1)) {evaluation_data[,i] = as.numeric(as.character(evaluation_data[,i]))}
```

Some columns were mostly blank.  These did not contribute well to the prediction.  I chose a feature
set that only included complete columns.  We also remove user name, timestamps and windows.  

Determine and display out feature set.

```{r}
feature_set <- colnames(training_data[colSums(is.na(training_data)) == 0])[-(1:7)]
model_data <- training_data[feature_set]
feature_set
```

We now have the model data built from our feature set.

```{r}
idx <- createDataPartition(y=model_data$classe, p=0.75, list=FALSE )
training <- model_data[idx,]
testing <- model_data[-idx,]
```

We now build 5 random forests with 150 trees each. We make use of parallel processing to build this
model. I found several examples of how to perform parallel processing with random forests in R, this
provided a great speedup.

```{r}
registerDoParallel()
x <- training[-ncol(training)]
y <- training$classe

rf <- foreach(ntree=rep(150, 6), .combine=randomForest::combine, .packages='randomForest') %dopar% {
randomForest(x, y, ntree=ntree) 
}
```

Provide error reports for both training and test data.
```{r}
predictions1 <- predict(rf, newdata=training)
confusionMatrix(predictions1,training$classe)


predictions2 <- predict(rf, newdata=testing)
confusionMatrix(predictions2,testing$classe)
```

Conclusions and Test Data Submit
--------------------------------

As can be seen from the confusion matrix this model is very accurate.  I did experiment with PCA 
and other models, but did not get as good of accuracy. Because my test data was around 99% 
accurate I expected nearly all of the submitted test cases to be correct.  It turned out they 
were all correct.

Prepare the submission. (using COURSERA provided code)

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}


x <- evaluation_data
x <- x[feature_set[feature_set!='classe']]
answers <- predict(rf, newdata=x)

answers

pml_write_files(answers)
```




































------------------------------------------------------------------------------------------------
```{r}
library(data.table)
library(caret)
library(ggplot2)
library(doMC)
library(knitr)
library(xtable)
library(randomForest)

registerDoMC(cores = 7)

## You should create a report describing how you built your model, how you used cross validation,
## what you think the expected out of sample error is, and why you made the choices you did.
## You will also use your prediction model to predict 20 different test cases.

## 1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML
## file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number
## of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages
## branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

## 2. You should also apply your machine learning algorithm to the 20 test cases available in the test data
## above. Please submit your predictions in appropriate format to the programming assignment for automated grading.
## See the programming assignment for additional details.

download.pml <- function() {
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv")
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv")
}

read.pml <- function(file) {
    fread(file, na.strings=c("#DIV/0!", ""))
}

build.report <- function() {
    knit2html("project.Rmd", "index.html")
}

raw.train <- read.pml("pml-training.csv")
raw.validation <- read.pml("pml-testing.csv")

set.seed(13)

## contains some NA values
na.cols <- raw.train[,sapply(.SD, function(x) any(is.na(x)))]

drop.columns <- function(x) {
    x[,!c("V1", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window"),with=F]
}

transform.features <- function(x) {
    x[,classe:=factor(classe)]
}

## try only columns that have values
training.features <- drop.columns(raw.train[,eval(names(which(na.cols == F))),with=F])

write.pml.predictions <- function(x) {
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

submit.prediction <- function(x, validation) {
    in.train <- createDataPartition(x$classe, p=.60, list=FALSE)
    train <- x[in.train[,1]]
    model.rf <- train(y=as.factor(train$classe), x=train[,!"classe",with=F], tuneGrid=data.frame(mtry=3), trControl=trainControl(method="none"), method="parRF")  
    write.pml.predictions(predict(model.rf, newdata=drop.columns(validation[,eval(names(which(na.cols == F))[-60]),with=F])))
}
```




























# Coursera Machine Learning Project

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Modeling

In order to provide consistent data to construct features, the first task was to sanitize the data.  The following steps where taken:

1.  Remove excel division error strings `#DIV/0!` and replace with `NA` values.
1.  Convert empty strings to `NA` values.

The primary code to preprocess the data is located [here](http://github.com/bpoweski/practical-machine-learning/blob/master/project.R).

There is additional code within the knitr markdown file [here](http://github.com/bpoweski/practical-machine-learning/blob/master/project.Rmd).

## Feature Selection

With a clean data set, the next task was to explore the data and determine what is likely useful information.  An important goal of any model to generalize well with unseen data.  Given this,

1.  Any features that contained NA values were removed.
1.  Several columns appeared to be entirely metadata.  These columns were dropped as any correlation that exists would likely be spurious and therefore cause the model to perform poorly.

The following fields removed included:

* the unlabled row index
* `user_name`
* `raw_timestamp_part_1`
* `raw_timestamp_part_2`
* `cvtd_timestamp`
* `new_window`
* `num_window`

## Cross Validation

Cross validation was achieved by splitting the training data into a test set and a training set using the following:

```{r cross_validate}
in.train <- createDataPartition(training.features$classe, p=.60, list=FALSE)

train <- training.features[in.train[,1]]
test <- training.features[-in.train[,1]]
```

The data was partioned by the `classe` variable to ensure the training set and test set contain examples of each class. 60% of the training data was allocated to the training set and the remainder for the validation set.

## Prediction

The random forest model was initially used to prediction.

```{r train, echo=F}
model.rf <- train(y=as.factor(train$classe), x=train[,!"classe",with=F], tuneGrid=data.frame(mtry=3), trControl=trainControl(method="none"), method="parRF")
```

```{r confusion_matrix, echo=F}
confusionMatrix(predict(model.rf, newdata=transform.features(test)), factor(test$classe))
```

It's estimated that the out of sample error would reflect the Kappa statistic of `r confusionMatrix(predict(model.rf, newdata=transform.features(test)), factor(test$classe))$overall["Kappa"]`.

### Variable Importance

```{r variable_importance, echo=F}
print(plot(varImp(model.rf, scale = FALSE)))
```

## Conclusion

The random forest algorithm appears to perform very well for predicting activities from accelerometers measurements.