---
title: "Loan Calculator Pitch"
author: "Frank D. Evans"
output: ioslides_presentation
---

## Amortization Calculator

A Loan calculator is a tool that helps a user estimate a monthly payment on an installment 
loan. The purpose of this tool is to make it easy for a user to make small adjustments 
to the amount of the loan, the length of time, and the interest rate--and receive instant 
feedback on the effect that those inputs will have on the monthly payment amount.  
  
The frequency of interest compounding can have a very profound effect, and this tool will 
help show the difference in a loan that compounds monthly vs yearly compounding.

## Demo of Underlying Function
```{r}
am_calc_cm <- function(amount, term_len, int_rate) {
    term_mon <- term_len * 12
    ir_per <- int_rate / (100 * 12)
    output_payment <- amount * 
        (((ir_per * (1+ir_per)^term_mon) / 
        (((1+ir_per)^term_mon) - 1)))
    output_payment
}
am_calc_cy <- function(amount, term_len, int_rate) {
    ir_per <- int_rate / 100
    output_payment <- amount * 
        (((ir_per * (1+ir_per)^term_len) / 
        (((1+ir_per)^term_len) - 1)))
    output_payment / 12
}
```
  
## Demo of Function Output

```{r}
amount <- 5000
term_len <- 7
int_rate <- 13.9
am_calc_cm(amount, term_len, int_rate)
am_calc_cy(amount, term_len, int_rate)
```
  
The function takes the term length parameter in years, and the interest rate in a 
percentage decimal number. Then the function itself converts the term length from 
years to months and the interest rate from a percentage to a straight floating point 
rate before making the final calculation of monthly payment amount.

## Effect of Interest

```{r echo=FALSE}
library(ggplot2)
v_int_rate <- seq(from=0.1, to=20.0, by=0.1)
scenarios <- data.frame(amount = rep(x=amount, times=length(v_int_rate)), 
                        term = rep(x=term_len, times=length(v_int_rate)), 
                        int_rate = v_int_rate)
scenarios$payment_cm <- am_calc_cm(scenarios$amount, scenarios$term, scenarios$int_rate)
scenarios$payment_cy <- am_calc_cy(scenarios$amount, scenarios$term, scenarios$int_rate)
ggplot(aes(x=int_rate), data=scenarios) + 
    geom_line(aes(y=payment_cm,), color='dark orange', lwd=2) +
    geom_line(aes(y=payment_cy), color='dark blue', lwd=2) +
    ggtitle('Demo of Interest Compounding Frequency ($5,000 - 7 year Example)') +
    xlab('Interest Rate (%) - Monthly [Orange] vs Yearly [Blue]') +
    ylab('Monthly Payment ($)')
```
  
From the example on the previous slide, we can chart the effect of a range of interest rates 
to see the difference in compounding cycles across the range on interst rates.

